**Evaluation of machine learning methods for 16S rRNA gene data**

### 2019-06-26 submission/TableS1.Rmd
update

\begin{center}
\captionof{table}{Model complexity: $n$ is the number of training samples, $f$ the number of features, $n_{trees}$ is the number of trees.}
\begin{tabular}{|l|l|}
\hline
\textbf{Machine learning model} & \textbf{Training complexity} \\ \hline
Logistic regression & $O(nf)$ \\ \hline
SVM with linear kernel & $O(nf)$  \\ \hline
SVM with radial basis kernel & $O(n^2f)$  \\ \hline
Decision tree & $O(n^2f)$  \\ \hline
Random forest & $O(n^2fn_{trees})$  \\ \hline
XGBoost & $O(nfn_{trees})?$  \\ \hline
\end{tabular}
\end{center}

>>>>

\begin{center}
\captionof{table}{Model complexity: $n$ is the number of training samples, $f$ the number of features, $n_{trees}$ is the number of trees.}
\begin{tabular}{|l|l|l|}
\hline
\textbf{Machine learning model} & \textbf{Training complexity} & \textbf{Hyperparameters} \\ \hline
Logistic regression & $O(nf)$ & $C$ \\ \hline
SVM with linear kernel & $O(nf)$ & $C$ \\ \hline
SVM with radial basis kernel & $O(n^2f)$ & $C$, $\sigma$ \\ \hline
Decision tree & $O(n^2f)$ & tree depth \\ \hline
Random forest & $O(n^2fn_{trees})$ & number of features \\ \hline
XGBoost & $O(nfn_{trees})?$  \\ \hline
\end{tabular}
\end{center}

### 2019-06-26 submission/manuscript.Rmd
update

######################################################################
source("../code/learning/load_AUROC_data.R")
```
\vspace{35mm}

@@

Machine learning (ML) modeling of the human microbiome has the potential to identify the microbial biomarkers and aid in diagnosis of many chronic diseases such as inflammatory bowel disease, diabetes and colorectal cancer. Progress has been made towards developing ML models that predict health outcomes from bacterial abundances, but rigourous ML models are scarce due to the flawed modeling methods that call into question the validity of developed ML models. Furthermore, using black box ML models has hindered the validation of microbial biomarkers. To overcome these challenges, we benchmarked seven different ML models that use fecal 16S rRNA sequences to predict the presence/absence of colorectal cancer (CRC) lesions (n=490 patients, 261 controls and 229 cases). To show the effect of model selection, we assessed the predictive performance, interpretability, and computational efficiency of the following models: L2-regularized logistic regression, L1 and L2 support vector machines (SVM) with linear and radial basis function kernels, a decision tree, random forest, and extreme gradient boosting (XGBoost). The random forest model was best at detecting CRC lesions with an AUROC of 0.695 but it was slow to train (101.3 h) and hard to interpret. Despite its simplicity, L2-regularized logistic regression followed random forest in predictive performance with an AUROC of 0.680, and it trained much faster (12 min). This study showed that we should choose ML models based on our expectations of predictive performance and interpretability as well as our  computational resources. It also established standards for modeling pipelines of microbiome-associated ML models.

@@

Prediction of health outcomes using ML is rapidly being adopted by human-associated microbiome studies. However, the developed ML so far are overoptimistic both in terms of predictive performance and validity. Without rigorous ML pipelines, we cannot trust ML models. Before we can speed up progress, we need to slow down, define and start using good ML practices.

@@

Machine learning (ML) methods are effective at recognizing and highlighting patterns in complex microbial datasets. Therefore, researchers have started to explore the utility of ML models that use microbiota associated biomarkers to predict human health and to understand the microbial ecology of diseases such as liver cirrhosis, colorectal cancer, inflammatory bowel diseases (IBD), obesity, type 2 diabetes and others [@zeller_potential_2014; @zackular_human_2014; @baxter_dna_2016; @baxter_microbiota-based_2016; @hale_shifts_2017; @pasolli_machine_2016; @sze_looking_2016; @walters_meta-analyses_2014; @vazquez-baeza_guiding_2018; @qin_alterations_2014; @geman_deep_2018].  However, currently the field's use of ML lacks clarity and consistency on which methods are used and how these methods are implemented [@thaiss_persistent_2016; dadkhah_gut_2019]. More notably, we commonly see ML practices such as using ML pipelines where there is no seperate held-out test dataset to evaluate model performance or reporting few or only the best outcomes of cross-validation. When there are seperate testing sets to evaluate model performance, there are large differences between cross-validation and testing performances as well as large confidence intervals of testing performances [@flemer_oral_2018; @baxter_microbiota-based_2016; @dai_multi-cohort_2018; @montassier_pretreatment_2016; @papa_non-invasive_2012; @mossotto_classification_2017; @ai_systematic_2017; @wong_quantitation_2017]. Moreover, there is a lack of discussion on why a particular ML model is utilized. Recently, there is a trend towards using more complex ML models such as random forest, extreme gradient boosting and neural networks without a discussion on if and how much model interpretibility is necessary for the study [@galkin_human_2018; @reiman_using_2017; @fioravanti_phylogenetic_2017; @ @geman_deep_2018]. The lack of transparency on modeling methodology and model selection negatively impact model reproducibility and reliability. We need to strive toward better machine learning practices by (1) implementing consistent and rigourous machine learning pipelines and (2) selecting ML models that reflect the goal of the study as it will inform our expectations of model accuracy, complexity, interpretibility and computational efficiency.

To showcase a reliable ML pipeline and to shed light on how much ML model selection can affect modeling results, we performed an empirical analysis comparing several different ML models using the same dataset and with a robust ML pipeline. We used a previously published colorectal cancer (CRC) study [@baxter_dna_2016] which had fecal 16S rRNA gene sequences from `r paste(nrow(shared))` patients. CRC is the third most common form of cancer worldwide in which the human-associated human microbiome is hypothesized to directly contribute to its development. We built ML models using fecal 16S rRNA gene sequences to predict patients with normal colons or patients with colonic tumors which are called screen relevant neoplasias (SRN). The study had `r paste(sum(sum(meta$dx=="normal")+sum(meta$Dx_Bin=="Adenoma")))` normal and `r paste(sum(sum(meta$dx=="cancer")+sum(meta$Dx_Bin=="adv Adenoma")))` SRN samples. We established modeling pipelines for L2-regularized logistic regression, L1 and L2 support vector machines (SVM) with linear and radial basis function kernels, a decision tree, random forest and XGBoost. Our ML pipeline utilized held-out test data to evaluate predictive performance and generalizability of each ML model. We used the area under the receiver operating characteristic curve (AUROC) as the predictive performance metric. The median test AUROC varied from `r paste(round(test_all[performance_index[1],3], 3))`  to `r paste(round(test_all[performance_index[7],3], 3))`. Random forest had the highest median AUROC for detecting SRN. Despite its simplicity, the L2-regularized logistic regression was second best in predictive performance. In terms of computational efficiency, L1 SVM with linear kernel trained the fastest (`r paste(round(traintime_df[traintime_index[1],2], 3))` hours, std ± `r paste(round(traintime_df[traintime_index[1],3], 3))`), while XGBoost took the longest (`r paste(round(traintime_df[traintime_index[7],2], 3))` hours, std ± `r paste(round(traintime_df[traintime_index[7],3], 3))`). We also found that depending on how the data is split to create a held-out test set, the AUROC values a ML model can vary up to `r paste(round(max((all %>% group_by(model) %>% summarise(difference=max(test_aucs)-min(test_aucs)))$difference), 3))` which highlighted the importance of performing many randomized modeling runs. Aside from evaluating generalization and predictive performances for each of these models, this study established standards for modeling pipelines of microbiome-associated machine learning models.

@@

We established a robust ML pipeline where we train and validate each of the seven models [Figure 1]. We randomly split the data into training/validation and test sets so that the training/validation set consisted of 80% of the full dataset while the test set was composed of the remaining data [Figure 1]. Since the cases are not uniformly represented in the data, the initial data-split was stratified to maintain the overall label distribution in both the training/validation and test sets. Training/validation set consisted of 393 patients (209 SRN), while the test set was composed of 97 patients (52 SRN). The training/validation data was used for training purposes and validation of hyperparameter selection (i.e. model selection), and the test set was used for evaluation purposes. Validation of hyperparameter selection was performed using repeated five-fold cross-validation on the training/validation set [Figure 1]. Similar to the initial data-split, five-fold cross-validation was also stratified to maintain the overall label distribution on the training and validation sets. We validated the cross-validation performances of each hyperparameter setting over the 100 randomizations and selected the best performing hyper-parameter setting in terms of AUROC to train the full training/validation dataset [Figures S1 and S2]. We then used the held-out test set to evaluate the prediction performance of each ML model. The data-split, hyperparameter selection, training and testing steps were repeated 100 times to get a reliable and robust reading of model prediction performance [Figure 1].

@@

We reported the testing AUROC values over 100 random splits of the full dataset which showed that depending on the data-split, the testing AUROC value showed great variability [Figure 1]. The testing AUROC values within each model varied `r paste(round(mean((all %>% group_by(model) %>% summarise(difference=max(test_aucs)-min(test_aucs)))$difference), 2))` on average across the seven models. For instance, the lowest AUROC value of the random forest model was `r paste(round(min((all %>% filter(model=="Random_Forest"))$test_aucs), 2))` whereas the highest was `r paste(round(max((all %>% filter(model=="Random_Forest"))$test_aucs), 2))`.

@@

Linear models trained faster than non-linear models. L2 logistic Rregression and L1 and L2 SVM with linear kernel had training times of `r paste(round(traintime_df[traintime_index[1],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[1],3], 2))`), `r paste(round(traintime_df[traintime_index[2],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[2],3],2))`), and `r paste(round(traintime_df[traintime_index[3],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[3],3], 2))`), respectively. Whereas, SVM with radial basis function kernel, decision tree, random forest and xgboost had training times of `r paste(round(traintime_df[traintime_index[4],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[4],3], 1))`), `r paste(round(traintime_df[traintime_index[5],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[5],3], 1))`), `r paste(round(traintime_df[traintime_index[6],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[6],3], 1))`) and `r paste(round(traintime_df[traintime_index[7],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[7],3], 1))`), respectively [Figure 4].

As the complexity of a ML model increase, its training times will increase as well (Table S1). When regularization and hyperparameter optimization (i.e. model selection) steps are added to the model pipeline, training times will increase further. The more hyperparameters a model has, the longer it will take to train. The computational burden that comes from model complexity can be reduced by parallelizing different segments of the ML pipeline. In this study we have parallelized each data-split which allowed 100 data-splits to be processed through the ML pipeline at the same time for each model. We can further parallelize the cross-validation step for each hyper-parameter setting.

@@

For L2-regularized logistic regression, L1 and L2 support vector machines (SVM) with linear and radial basis function kernels we tuned the __cost__ hyperparameter which determines the regularization strength where smaller values specify stronger regularization. For SVM with radial basis function kernel we also tuned __sigma__ hyperparameter which determines the reach of a single training instance where for a high value of sigma, the SVM decision boundary will be dependent on the points that are closest to the decision boundary. For the decision tree model, we tuned the __depth of the tree__ where deeper the tree, the more splits it has. For random forest, we tuned the __number of features__ to consider when looking for the best tree split. For xgboost, we tuned for __learning rate__ and the __fraction of samples__ to be used for fitting the individual base learners.

Hyperparameter selection is very important. We start with a granular grid search by then narrow and refine. Range of the grid depends on the problem. We can use hyper-band to help us with our hyperparameter selection. If not a full grid search is done, there could be a lot of variability in testing performance.
Models were trained using the machine learning wrapper caret package (v.6.0.81) in R (v.3.5.0).

>>>>

######################################################################
source("../code/learning/load_AUROC_data.R")
```

@@

Machine learning (ML) modeling of the human microbiome has the potential to identify the microbial biomarkers and aid in diagnosis of many chronic diseases such as inflammatory bowel disease, diabetes and colorectal cancer. Progress has been made towards developing ML models that predict health outcomes from bacterial abundances, but rigourous ML models are scarce due to the flawed modeling methods that call into question the validity of developed ML models. Furthermore, the use of black box ML models has hindered the validation of microbial biomarkers. To overcome these challenges, we benchmarked seven different ML models that use fecal 16S rRNA sequences to predict the presence/absence of colorectal cancer (CRC) lesions (n=490 patients, 261 controls and 229 cases). To show the effect of model selection, we assessed the predictive performance, interpretability, and computational efficiency of the following models: L2-regularized logistic regression, L1 and L2 support vector machines (SVM) with linear and radial basis function kernels, a decision tree, random forest, and extreme gradient boosting (XGBoost). The random forest model was best at detecting CRC lesions with an AUROC of 0.695 but it was slow to train (101.3 h) and hard to interpret. Despite its simplicity, L2-regularized logistic regression followed random forest in predictive performance with an AUROC of 0.680, and it trained much faster (12 min). This study showed that we should choose ML models based on our expectations of predictive performance, interpretability and our computational resources. It also established standards for modeling pipelines of microbiome-associated ML models.

@@

Prediction of health outcomes using ML is rapidly being adopted by human-associated microbiome studies. However, the developed ML so far are overoptimistic in terms of predictive performance and validity. Without rigorous ML pipelines, we cannot trust ML models. Before we can speed up progress, we need to slow down, define and start using good ML practices.

@@

Machine learning (ML) methods are effective at recognizing and highlighting patterns in complex microbial datasets. Therefore, researchers have started to explore the utility of ML models that use microbiota associated biomarkers to predict human health and to understand the microbial ecology of diseases such as liver cirrhosis, colorectal cancer, inflammatory bowel diseases (IBD), obesity, type 2 diabetes and others [@zeller_potential_2014; @zackular_human_2014; @baxter_dna_2016; @baxter_microbiota-based_2016; @hale_shifts_2017; @pasolli_machine_2016; @sze_looking_2016; @walters_meta-analyses_2014; @vazquez-baeza_guiding_2018; @qin_alterations_2014; @geman_deep_2018].  However, currently the field's use of ML lacks clarity and consistency on which methods are used and how these methods are implemented [@thaiss_persistent_2016; dadkhah_gut_2019]. More notably, we commonly see ML practices such as using ML pipelines where there is no seperate held-out test dataset to evaluate model performance or reporting few or only the best outcomes of cross-validation. Even when there are seperate testing sets to evaluate model performance, we observe large differences between cross-validation and testing performances that indicate overfitting as well as large confidence intervals for testing performances [@flemer_oral_2018; @baxter_microbiota-based_2016; @dai_multi-cohort_2018; @montassier_pretreatment_2016; @papa_non-invasive_2012; @mossotto_classification_2017; @ai_systematic_2017; @wong_quantitation_2017]. Moreover, there is a lack of discussion on why a particular ML model is utilized. Recently, there is a trend towards using more complex ML models such as random forest, extreme gradient boosting and neural networks without a discussion on if and how much model interpretibility is necessary for the study [@galkin_human_2018; @reiman_using_2017; @fioravanti_phylogenetic_2017; @ @geman_deep_2018]. Black box machine learning models require posthoc explanations to determine the feature importances in making a prediction. These explanations can be misleading and at times unreliable when making high-stake decisions about someone's health [@rudin_please_2018]. The lack of transparency on model selection and interpretation as well as flawed modeling methods negatively impact model  validity and reproducibiliyu. We need to strive toward better machine learning practices by (1) implementing rigourous machine learning pipelines and (2) selecting ML models that reflect the goal of the study as it will inform our expectations of model accuracy, complexity, interpretibility and computational efficiency.

To showcase a rigorous ML pipeline and to shed light on how much ML model selection can affect modeling results, we performed an empirical analysis comparing several different ML models using the same dataset and the same ML pipeline. We used a previously published colorectal cancer (CRC) study [@baxter_dna_2016] which had fecal 16S rRNA gene sequences from `r paste(nrow(shared))` patients. CRC is a type of cancer which the human-associated human microbiome is hypothesized to directly contribute to its development therefore fecal 16S rRNA gene sequences can be used to detect CRC. We built ML models using fecal 16S rRNA gene sequences to predict healthy patients versus patients with colorectal lesions that were identified by colonoscopy as screen relevant neoplasias (SRN). The study had `r paste(sum(sum(meta$dx=="normal")+sum(meta$Dx_Bin=="Adenoma")))` normal and `r paste(sum(sum(meta$dx=="cancer")+sum(meta$Dx_Bin=="adv Adenoma")))` SRN samples. We established modeling pipelines for L2-regularized logistic regression, L1 and L2 support vector machines (SVM) with linear and radial basis function kernels, a decision tree, random forest and XGBoost. Our ML pipeline utilized held-out test data to evaluate predictive performance and generalizability of each ML model. We used the area under the receiver operating characteristic curve (AUROC) as the predictive performance metric. The median test AUROC varied from `r paste(round(test_all[performance_index[1],3], 3))`  to `r paste(round(test_all[performance_index[7],3], 3))`. Random forest had the highest median AUROC for detecting SRN. Despite its simplicity, the L2-regularized logistic regression was second best in predictive performance. In terms of computational efficiency, L1 SVM with linear kernel trained the fastest (`r paste(round(traintime_df[traintime_index[1],2], 3))` hours, std ± `r paste(round(traintime_df[traintime_index[1],3], 3))`), while XGBoost took the longest (`r paste(round(traintime_df[traintime_index[7],2], 3))` hours, std ± `r paste(round(traintime_df[traintime_index[7],3], 3))`). We also found that depending on how the data is split to create a held-out test set, the AUROC values of a ML model can vary up to `r paste(round(max((all %>% group_by(model) %>% summarise(difference=max(test_aucs)-min(test_aucs)))$difference), 3))` which highlighted the importance of performing many randomized modeling runs. This study established standards for microbiome-associated ML models and underscored the importance of model selection.

@@

We established a ML pipeline where we train and validate each of the seven models [Figure 1]. We randomly split the data into training/validation and test sets so that the training/validation set consisted of 80% of the full dataset while the test set was composed of the remaining data [Figure 1]. Since the cases are not uniformly represented in the data, the initial data-split was stratified to maintain the overall label distribution in both the training/validation and test sets. Training/validation set consisted of 393 patients (209 SRN), while the test set was composed of 97 patients (52 SRN). The training/validation data was used for training purposes and validation of hyperparameter selection (i.e. model selection), and the test set was used for evaluation purposes. Validation of hyperparameter selection was performed using repeated five-fold cross-validation on the training/validation set [Figure 1]. Similar to the initial data-split, five-fold cross-validation was also stratified to maintain the overall label distribution on the training and validation sets. We validated the cross-validation performances of each hyperparameter setting over the 100 randomizations and selected the best performing hyper-parameter setting in terms of AUROC to train the full training/validation dataset [Figures S1 and S2]. We then used the held-out test set to evaluate the prediction performance of each ML model. The data-split, hyperparameter selection, training and testing steps were repeated 100 times to get a reliable and robust reading of model performance [Figure 1].

@@

We reported the testing AUROC values over 100 randomizations of the initial data-split. The results showed that depending on the data-split, the testing AUROC value showed great variability [Figure 2]. The testing AUROC values within each model varied `r paste(round(mean((all %>% group_by(model) %>% summarise(difference=max(test_aucs)-min(test_aucs)))$difference), 2))` on average across the seven models. For instance, the lowest AUROC value of the random forest model was `r paste(round(min((all %>% filter(model=="Random_Forest"))$test_aucs), 2))` whereas the highest was `r paste(round(max((all %>% filter(model=="Random_Forest"))$test_aucs), 2))`.

@@

As the complexity of a ML model increased [Table S1], its training times increased as well [Figure 5]. Additionally, the more hyperparameters a model had [Table S1], the longer it took to train [Figure 5]. Linear models trained faster than non-linear models. L2 logistic Rregression and L1 and L2 SVM with linear kernel had training times of `r paste(round(traintime_df[traintime_index[1],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[1],3], 2))`), `r paste(round(traintime_df[traintime_index[2],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[2],3],2))`), and `r paste(round(traintime_df[traintime_index[3],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[3],3], 2))`), respectively. Whereas, SVM with radial basis function kernel, decision tree, random forest and xgboost had training times of `r paste(round(traintime_df[traintime_index[4],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[4],3], 1))`), `r paste(round(traintime_df[traintime_index[5],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[5],3], 1))`), `r paste(round(traintime_df[traintime_index[6],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[6],3], 1))`) and `r paste(round(traintime_df[traintime_index[7],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[7],3], 1))`), respectively [Figure 5].

@@

Models were trained using the machine learning wrapper caret package (v.6.0.81) in R (v.3.5.0). Within the caret package, we have made modifications to L2-regularized SVM with linear kernel function __svmLinear3__ and developed a L1-regularized SVM with linear kernel function __svmLinear4__ to calculate decision values instead of predicted probabilities. These changes are available at https://github.com/SchlossLab/Topcuoglu_ML_XXXX_2019/.

For L2-regularized logistic regression, L1 and L2 support vector machines (SVM) with linear and radial basis function kernels we tuned the __cost__ hyperparameter which determines the regularization strength where smaller values specify stronger regularization. For SVM with radial basis function kernel we also tuned __sigma__ hyperparameter which determines the reach of a single training instance where for a high value of sigma, the SVM decision boundary will be dependent on the points that are closest to the decision boundary. For the decision tree model, we tuned the __depth of the tree__ where deeper the tree, the more splits it has. For random forest, we tuned the __number of features__ to consider when looking for the best tree split. For XGBoost, we tuned for __learning rate__ and the __fraction of samples__ to be used for fitting the individual base learners.For hyperparameter selection, we started with a granular grid search. Then we narrowed and fine-tuned the range of each hyperparameter. The range of the grid depends on the ML task and ML model. A full grid search needs to be performed to avoid variability in testing performance. We can use hyper-band to help us with our hyperparameter selection [@li_hyperband:_2016].

The computational burden during model training due to model complexity was reduced by parallelizing segments of the ML pipeline. In this study we have parallelized each data-split which allowed 100 data-splits to be processed through the ML pipeline at the same time for each model. We can further parallelize the cross-validation step for each hyper-parameter setting.
